Can We Crunch the Social Graph on the Cheap?

### 2011-01-24 doc/hotpar2011/softxmt-hotpar2011.tex
reworked evaluation

\documentclass{acm_proc_article-sp}
\usepackage{url}
\begin{document}

@@

\numberofauthors{7}
\author{
(authors)
}

@@

@@

\section{Current Implementation and Results}

\todo{Begin eval insert}
\input{evaluation}
\todo{End eval insert}

>>>>

\documentclass{acm_proc_article-sp}
\usepackage{url}
\usepackage{fancyvrb}
\usepackage{alltt}
\begin{document}

@@

\numberofauthors{7}
\author{
Jacob Nelson, Brandon Myers, Andrew Hunter, Dan Grossman, Luis Ceze\\
University of Washington\\
\email{\{nelson, bdmyers, ahh, djg, luisceze\}@cs.washington.edu}\\ \\
Simon Kahan \\ 
Pacific Northwest National Lab\\
\email{\{skahan\}@cs.washington.edu}
}

@@

Graph analytics are very important.

Irregular applications present a challenge for parallel machines:
obtaining performance is not straightforward.

Multithreading is a technique that has been successfully used to
extract performance from these problems.

Our goal is to build a system with XMT-like properties, implemented in
software with commodity procesors

In this paper, we describe a runtime for latency tolerance. Our
analysis shows the feasibility of our approach. We end with our plans
for adding synchronization support and scaling up to multiple nodes.

@@

\section{Evaluation}

To evaluate the feasibility of our proposal, we ran pointer chasing
benchmarks on a single-node implementation of our runtime. These
pointer chasing experiments are intended to model a particular
``wost-case'' behavior of irregular applications, where each memory
reference causes a cache miss.

We ran these experiments on a Dell PowerEdge R410 with two Xeon X5650
processors and 24GB of RAM, with hyperthreading disabled. These
processors use a NUMA memory architecture, where each processor has
its own integrated memory controller and DIMMs; references to other
processors are carried over Intel's cache-coherent QuickPath
Interconnect (QPI).

Our evaulation consists of three parts. First, we describe the
performance of the test machine's memory system to use as a
baseline. Then, we demonstrate that the same performance is achievable
using our runtime system: we measure pointer chasing performance and
the effects of cache pressure due to the number of contexts
required. Finally, we simulate the effects of fetching memory over a
network by adding delays to our runtime.

\subsection{Memory system performance}

We first characterize the performance of our test machine's memory
system. We measured three parameters: the maximum random reference
rate that can be issued by the cores in one processor, the maximum random
reference rate that can be serviced by one processor's memory controller,
and the maximum random reference rate that can be carried over the QPI
link and serviced by another processor's memory controller.

To find the maximum random reference rate that one processor's cores
can issue, we ran pointer chasing code following the model of
Figure~\ref{fig:pointernocoro}. Each core issues $n$ list traversals
in a loop; we call $n * \textit{number of cores}$ the number of
{\em concurrent offered references}, since the memory system may not
be able to satisfy them all in parallel. Since our goal is to find a
baseline for evaluating our coroutine library, we depend on the
processor's ILP for memory concurrency, rather than our coroutine
library.

The lists were sized to exceed the last level cache and were laid out
randomly with pointers spaced at a cache line granularity, maximizing
the probability of each reference being a miss. We allocated the lists
on 1GB huge pages to minimize TLB refill overhead. The lists were
allocated in the memory attached to the processor containing the cores
doing the traversal. We ran 10 traversals and recorded the maximum
rate.

\begin{figure}[h]
	\begin{center}
		\includegraphics[width=0.5\textwidth]{figures/multi-listwalk-totalconc-edited.pdf}
	\end{center}
	\caption{Throughput of listwalk versus total number of
          concurrent references offered by the cores. Data points for
          running 1 core are shown as circles, and data points for
          running 6 cores are shown as diamonds. 
%Notice that the throughput for a single core levels out at 10
%concurrent references. For 4 to 6 cores, the throughput levels off at
%around 36 references, which seems to be the most memory concurrency a
%processor can handle.
        }
	\label{fig:listwalk-totalconc}
\end{figure}

Figure~\ref{fig:listwalk-totalconc} shows the result. Each point represents
the maximum rate pointers are traversed for a given number of
concurrent offered references. We see a maximum rate of 277 million
references per second; this is achieved when the number of offered
references is 42. Note that a single core cannot support this level of
memory concurrency; the maximum reference rate for a single core is
107M. We believe this limit is due to the core's {\em line fill
  buffer} \todo{cite nehalem perf docs}, which limits the core to 10
concurrent L2 misses.

The memory controller in the processor has more bandwidth than its
cores can saturate. To measure the memory controller's maximum random
reference rate, we extended the previous experiment so that cores in
both processors were traversing lists allocated in the first processor's
memory. With this configuration, we observed a maximum rate of 360
million references per second. We believe this difference is due to
the configuration of the processor's {\em Global Queue} \todo{cite nehalem
  docs}, which limits the number of concurrently-executing read misses
from a processor's cores to its local last-level cache and memory to 32.

\begin{figure}[h]
  \begin{center}
    \includegraphics[width=0.5\textwidth]{figures/qpi_bw-edited.pdf}
  \end{center}
  \caption{Pointer chasing in a remote processor's memory.}
  \label{fig:listwalk-qpi}
\end{figure}

Finally, we measured the rate at which the cores can make requests
over the QPI link. For this experiment, we allocated the lists in the
second processor, and ran the traversals on the first
processor. Figure~\ref{listwalk-qpi} shows the result. We found that use
of the QPI link limits us to 175 million references per second.

\begin{figure}
\begin{alltt}
  while (count-- > 0) \{
    list1 = list1->next;
    list2 = list2->next;
    \ldots
    list\(n\) = list\(n\)->next;
  \}
\end{alltt}
\caption{Pseudocode for pointer chasing without coroutines.}
\label{fig:pointernocoro}
\end{figure}

\subsection{Runtime performance}

To characterize the performance of our runtime, we investigated three
effects: the maximum reference rate using coroutines
to obtain memory concurrency, the effects of cache pressure from the
coroutines' context storage, and the maximum reference rate obtainable
using a simulated network delay.

\begin{figure}
\begin{alltt}
  while (count-- > 0) \{
     prefetch(&(list1->next));
     prefetch(&(list2->next));
     \ldots
     prefetch(&(list\(n\)->next));
     switch();
     list1 = read(&(list1->next));
     list2 = read(&(list2->next));
     \ldots
     list\(n\) = read(&(list\(n\)->next));
 \}
\end{alltt}
\caption{Pseudocode for pointer chasing using coroutines.}
\label{fig:pointercoro}
\end{figure}

To find the maximum reference rate obtainable using our coroutine
library, we modified our pointer chasing benchmark as shown in
figure~\ref{fig:pointercoro}. Now there are two sources of offered
concurrent references: the ILP exploited by the processor, and the
memory concurrency enabled by prefetching and switching to a new
coroutine. As with our first experiment, we allocated the lists in the
same processor as the cores doing the traversal.

\begin{figure}[h]
  \begin{center}
    \includegraphics[width=0.5\textwidth]{figures/multi-green-edited.pdf}
  \end{center}
  \caption{Pointer chasing with coroutines, with one reference per
    coroutine.}
  \label{fig:multi-green}
\end{figure}

Figure~\ref{fig:multi-green} shows the result. For this
experiment, we limited memory concurrency to one concurrent miss per
coroutine, so the only source of memory concurrency is the use of
coroutines. We are able to obtain a rate of 275 million references per
second with 48 concurrent misses, or 8 coroutines per core. These
results are very close to the ILP-only experiment.

We observe a gradual decrease in reference rate once the number of
concurrent references per core exceeds 10; we believe this is due to
later prefetches squashing earlier ones in the line fill buffer. We
also observe many samples below the maximum rate even with enough
memory concurrency; this is harder to explain, but we suspect it is
due to resource contention in the cores' pipelines once the memory
system is full of requests.

\begin{figure}
\begin{alltt}
  while (count-- > 0) \{
    prefetch(&(list1->next));
    switch();
    list1 = read(&(list1->next));
    for( i in 1 to num_local_updates ) \{
      local->data++;
      local = local->next;
   \}
  \}
\end{alltt}
\caption{Pseudocode for pointer chasing with local updates.}
\label{pointerupdate}
\end{figure}

The stacks for the coroutines are stored in the data cache, where they
will compete for space with an application's data. To characterize the
effects of this cache pressure, we modified our list chasing benchmark
to include random updates to a per-coroutine local data structure that
is small enough to fit in cache. Figure~\ref{pointerupdate} shows the
general idea.

\begin{figure}[h]
  \begin{center}
    \includegraphics[width=0.5\textwidth]{figures/cache-pressure-edited.pdf}
  \end{center}
  \caption{Cache pressure with coroutines.}
  \label{fig:cache-pressure}
\end{figure}

We ran this code on six cores with all data allocated in the same
processor's memory. We varied the size of the local working set from 2KB
to 4MB, and ran both 0 and 32 local updates for each pointer
traversal. We measured the overall reference rate, including both the

Figure~\ref{fig:cache-pressure} shows the result. On the left of the
figure, we see the list-traversal-only case, where we allocate
per-coroutine space, but make no references to it. As we would expect,
the reference rates grow with the number of coroutines just as in
Figure~\ref{fig:multi-green}. On the right of the figure, we see the
case with 32 updates for each list traversal. While performance
decreases for large working set sizes and large number of coroutines,
we see speedup for up to 128KB working set size with 8 coroutines per
core.

Finally, to simulate the performance of pointer chasing in a
multi-node system, we modified the scheduler of our coroutine library
to include a delay before a coroutine can be rescheduled, imitating
the network transit delay even though we are referencing local
memory. Recently published measurements of Infiniband latencies
\todo{cite} suggest we might expect 1.1 $\mu$s source-to-destination
latencies between two computers connected back-to-back; we assume a
switch transit delay of 400 ns in each direction and thus estimate a 3
$\mu$s round-trip delay. In this experiment, once a coroutine has
yielded, we use the core's timestamp counter to ensure it does not get
rescheduled for 3 $\mu$s.

\begin{figure}[h]
  \begin{center}
    \includegraphics[width=0.5\textwidth]{figures/multi-green-delay7800-edited.pdf}
  \end{center}
  \caption{Pointer chasing with simulated network delay.}
  \label{fig:network-delay}
\end{figure}

We ran these experiments with two sources of memory concurrency: each
core ran up to 250 coroutines, and each coroutine made between 1 and 4
concurrent references. The results are shown in
Figure~\ref{fig:network-delay}. We are still able to reach a
near-maximum rate of 273 million references per second, but at least 3
concurrent references per coroutine are required. Since a single
reference per coroutine achieves 236 million references per
second. Any network interface would be connected through a QPI link,
which we measured to have a bandwidth limit of 175 million references
per second, our runtime appears to have the capacity to chase pointers
in a remote memory.

\todo{may want to rerun with memory on remote core to incorporate QPI limit?}

%\todo{Begin eval insert}
%\input{evaluation}
%\todo{End eval insert}
